Global optimization algorithm that uses decision trees to find local extrema \thanks{The work was supported by the Ministry of Science and Higher Education of the Russian Federation (project no. 0729-2020-0055), and by the Research and Education Mathematical Center (project no. 075-02-2022-883).}

### 2023-01-30 2023/problem-info/artricle.tex
Поправлен английский текст.

\title{Global optimization algorithm using decision trees to find local extrema \thanks{The work was supported by the oneAPI Center of Excellence, by the Ministry of Science and Higher Education of the Russian Federation (project no. 0729-2020-0055), and by the Research and Education Mathematical Center (project no.  075-02-2021-1394).}}

@@

\begin{small}
\textbf{Abstract:} To solve multidimensional global optimization problems, local methods can be used to locally refine and accelerate the convergence of the entire algorithm. In order to identify areas of attraction of local minima, decision trees were used in this work. For the objective function of the optimization problem, it is only assumed that it satisfies the Lipschitz condition, while the constant is considered unknown. To confirm the theory, a number of experiments were given. They demonstrate that when using this approach on a series of test problems, it is possible to achieve a significant acceleration in the number of iterations of the algorithm.

@@

%%%%%% 
%%%%%% Информация на русском языке
%%%%%%
\emptythanks
\title{Алгоритм глобальной оптимизации, использующий деревья решений для выявления локальных экстремумов\thanks{Работа выполнена при финансовой поддержке Министерства науки и высшего образования РФ (проект № 0729-2020-0055) и научно-образовательного математического центра “Математика технологий будущего” (соглашение № 075-02-2021-1394).}}

@@

\begin{small}
\textbf{Аннотация:} Для решения задач многомерной глобальной оптимизации можно использовать локальные методы для локального уточнения и ускорения сходимости всего адгоритма. С целью выявления областей притяжения локальных минимумов, в данной работе применялись деревья решений. Для целевой функции задачи оптимизации предполагается только то, что она удовлетворяет условию Липшица, при этом константа считается неизвестной. Для подтверждения теории был приведен ряд экспериментов. Они демонстрируют, что при использовании данного подхода на серии тестовых задач можно добиться значительного ускорения по числу итераций алгоритма.

@@

\paragraph{Дмитрий Игоревич Силенко} --- окончил бакалавриат института информационных технологий, математики и механики Нижегородского государственного университета им. Н. И. Лобачевского в 2021 году. В настоящее время является магистрантом института информационных технологий, математики и механики (направление - прикладная математика и информатика). Область научных интересов включает методы глобальной и локальной оптимизации, а также технологии параллельных вычислений и параллельные алгоритмы. e-mail: \url{rotor12587@mail.ru}. Является автором или соавтором нескольких научных работ в указанных областях.

@@

\paragraph{Ilya~G.~Lebedev} --- Head of the Laboratory of Supercomputer Technologies and High-Performance Computing, Department of Software and Supercomputer Technologies, Institute of Information Technology, Mathematics and Mechanics, Nizhny Novgorod State University. N.I. Lobachevsky. e-mail: \url{ilya.lebedev@itmm.unn.ru}. His research interests include algorithms of global and local optimization, parallel algorithms, GPU programming, CUDA. He has authored or coauthored more than 30 papers in these areas.

\paragraph{D. I. Silenko} --- Graduated the Institute of Information Technology, Mathematics and Mechanics of the Lobachevsky State University of Nizhni Novgorod in 2021. He is currently a master student of the Institute of Information Technology, Mathematics and Mechanics. His research interests are algorithms of global and local optimization, parallel algorithms and parallel technologies. e-mail: \url{rotor12587@mail.ru}. He has authored or coauthored several papers in these areas.

>>>>

\title{Global optimization algorithm that uses decision trees to find local extrema \thanks{The work was supported by the Ministry of Science and Higher Education of the Russian Federation (project no. 0729-2020-0055), and by the Research and Education Mathematical Center (project no.  075-02-2022-883).}}

@@

\begin{small}
\textbf{Abstract:} The paper considers the algorithms for solving the multidimensional global optimization problems using decision tree to reveal the attraction regions of the local minima. We suppose, that the target function is defined as a “black box” and satisfied Lipschitz condition with unknown constant. We propose a method for selecting the local extrema neighborhood of the target function based on analysis of accumulated search information using machine learning methods. This allows us to make a decision to run a local method, which can speed up the convergence of the algorithm. The proposition was confirmed by the results of numerical experiments demonstrating the speedup when solving a series of test problems.

@@

%%%%%% 
%%%%%% Информация на русском языке
%%%%%%
\emptythanks
\title{Алгоритм глобальной оптимизации, использующий деревья решений для выявления локальных экстремумов\thanks{Работа выполнена при финансовой поддержке Министерства науки и высшего образования РФ (проект № 0729-2020-0055) и научно-образовательного математического центра “Математика технологий будущего” (соглашение № 075-02-2022-883).}}

@@

\begin{small}
\textbf{Аннотация:} В работе рассматривается решение задач многомерной глобальной оптимизации с применением деревьев решений для выявления областей притяжения локальных минимумов. Предполагается, что целевая функция задачи задана как «черный ящик» и удовлетворяет условию Липшица с априори неизвестной константой. Мы предлагаем способ выделения окрестностей локальных экстремумов целевой функции на основе анализа накопленной поисковой информации средствами машинного обучения. Это позволяет принять решение о запуске локального метода для уточнения значения функции в локальном минимуме, что может ускорить сходимость алгоритма. Выдвинутое предположение подтверждается результатами вычислительных экспериментов, демонстрирующих ускорение при решении серии тестовых задач.

@@

\paragraph{Дмитрий Игоревич Силенко} --- окончил бакалавриат института информационных технологий, математики и механики Нижегородского государственного университета им. Н. И. Лобачевского в 2021 году. В настоящее время является магистрантом института информационных технологий, математики и механики (направление - прикладная математика и информатика). Область научных интересов включает методы глобальной и локальной оптимизации, а также технологии параллельных вычислений и параллельные алгоритмы. e-mail: \url{silenko@itmm.unn.ru}. Является соавтором нескольких научных работ в указанных областях.

@@

\paragraph{Ilya~G.~Lebedev} --- Head of the Laboratory of Supercomputer Technologies and High-Performance Computing, Department of Mathematical Software and Supercomputing Technologies, Institute of Information Technology, Mathematics and Mechanics, Nizhny Novgorod State University. N.I. Lobachevsky. e-mail: \url{ilya.lebedev@itmm.unn.ru}. His research interests include algorithms of global and local optimization, parallel algorithms, GPU programming, CUDA. He is the author or coauthor more than 30 papers in these areas.

\paragraph{D. I. Silenko} --- Graduated the Institute of Information Technology, Mathematics and Mechanics of the Lobachevsky State University of Nizhni Novgorod in 2021. He is currently a master student of the Institute of Information Technology, Mathematics and Mechanics. His research interests are algorithms of global and local optimization, parallel algorithms and parallel computing. e-mail: \url{silenko@itmm.unn.ru}. He is coauthor of the several papers in these areas.
